# Unraveling Notion of Dependence through Copulas {#sec-simple}

```{r}
#| label: setup
#| include: false

# Load packages
library(tidyverse)
library(gt)

# Set default ggplot theme for document
theme_set(theme_classic())
# If using kableExtra tables, print blank cells instead of `NA`
options(knitr.kable.NA = "")

# Load data
load("data/temp_wnba.RData")
```

Make it accessible by starting with a motivating example, then proceed towards more rigorousness gradually until challenges for discrete case are covered.


```{r}
#| label: fig-wnba-ht
#| fig-cap: "Distribution of heights of WNBA players in the 2024 season."

# Use Freedman-Diaconus rule to set binwidth
ht_bw <- 2 * IQR(wnba$height) / nrow(wnba)^(1/3)

# Create histogram of height faceted by player position
ggplot(wnba, aes(height)) +
  geom_histogram(binwidth = ht_bw) +
  labs(x = "Height (in)",
       y = "Count",
       caption = "Source: https://www.espn.com/wnba/stats/player") +
  theme_bw()
```


## Tables

Your tables should be publication quality. Consider using [**gt**](https://gt.rstudio.com/articles/gt.html) [@gt] or  [**kableExtra**](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_pdf.pdf) [@kableExtra] to customize your tables. The [**gtsummary**](https://www.danieldsjoberg.com/gtsummary/) package [@gtsummary] may also come in handy.

@tbl-ht-by-pos shows the average heights of WNBA players by position. 

```{r}
#| label: tbl-ht-by-pos
#| tbl-cap: "Average WNBA player height by position."

wnba |> 
  group_by(position) |> 
  summarize(mean_ht = mean(height)) |> 
  gt() |> 
  cols_label(
    position = "Position",
    mean_ht = "Average height (in)"
  ) |> 
  fmt_number(decimals = 1)
```


<!-- 
The following section should be included at the end of each chapter that contains code.

Note that this will include code from code chunks even if `eval` is set to `false`.
-->

## @sec-simple Code

The following code was used to create @sec-simple.

### Code within chapter
```{r}
#| echo: true
#| eval: false
#| ref-label: !expr knitr::all_labels()
```

### Code sourced from external scripts
```{r}
#| echo: true
#| eval: false
{{< include src/temp01-clean-wnba.R >}}
```
